Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy

### 2024-08-27 paper/analysis.tex
Section 4 draft.

As discussed in detail in \cite{2008ApJ...684..287I}, the photometric metallicity is best constrained for blue main-sequence stars with
$g-i<1$ (equivalently, $0.2 < g − r < 0.6$).
Therefore, we expect the best statistical behavior (smallest bias and scatter, $\chi^2 \sim 1$) in this color range, which is consistent with
the behavior seen in Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH}. An effective depth limit is about
$u=25$ (about 1 mag brighter than the 5-$\sigma$ depth in the $u$ band for coadded LSST photometry,  \citealt{2022ApJS..258....1B}); 
at fainter magnitudes the scatter between true and estimated values
for both $M_r$ and $[Fe/H]$ rapidly increases.

@@

We first analyze results for a subsample of stars that have $0.2 < g − r < 0.6$, a color range where photometric metallicity estimator is best constrained,
and a signal-to-noise ratio for Gaia's trigonometric
parallax measurement above 10.  We also required $u<21$ but that requirement is less stringent than the implied flux limit due to
Gaia's signal-to-noise ratio limit (approximately, $u<19$). There are about 18,000 stars in this subsample.
We find that their $M_r$ is estimated using SDSS photometry with a bias of 0.06 mag and root-mean-square scatter of 0.30 mag, compared
to Gaia's measurements. Equivalently, photometric distances  are estimated with a scatter of 15\% and a bias of 3\%.

>>>>

As discussed in detail in \cite{2008ApJ...684..287I}, the photometric metallicity is best constrained for blue main-sequence stars with
$g-i<1$ (equivalently, $0.2 < g - r < 0.6$).
Therefore, we expect the best statistical behavior (smallest bias and scatter, $\chi^2 \sim 1$) in this color range, which is consistent with
the behavior seen in Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH}. An effective depth limit is about
$u=25$ (about 1 mag brighter than the 5-$\sigma$ depth in the $u$ band for coadded LSST photometry,  \citealt{2022ApJS..258....1B}); 
at fainter magnitudes the scatter between true and estimated values
for both $M_r$ and $[Fe/H]$ rapidly increases.

@@

We first analyze results for a subsample of stars that have $0.2 < g - r < 0.6$, a color range where photometric metallicity estimator is best constrained,
and a signal-to-noise ratio for Gaia's trigonometric
parallax measurement above 10.  We also required $u<21$ but that requirement is less stringent than the implied flux limit due to
Gaia's signal-to-noise ratio limit (approximately, $u<19$). There are about 18,000 stars in this subsample.
We find that their $M_r$ is estimated using SDSS photometry with a bias of 0.06 mag and root-mean-square scatter of 0.30 mag, compared
to Gaia's measurements. Equivalently, photometric distances  are estimated with a scatter of 15\% and a bias of 3\%.

### 2024-08-27 paper/pipeline.tex
comments from coauthors
Section 4 draft.

One of recent tools developed to efficiently loop across the sky patches and perform some per-patch calculations
is the  Large Survey Database...

>>>>

One of the recent tools developed to efficiently loop across the sky patches and perform some per-patch calculations is the \href{https://lsdb.readthedocs.io/}{Large Survey Database}\footnote{https://lsdb.readthedocs.io/}.

The combination of the HiPSCat and LSDB workflow are particularly efficient and useful for a science case tackled by the PhotoD. HiPSCat effectively partitions the sky in pixels based on the density of the sources, therefore creating pixels that contain roughly the same number of stars. This strategy is appropriate both scientifically, as the resolution of the map (area subtended by the pixel) depends on the number of stars within (e.g. smaller pixels and higher resolution in the galactic plane and lower in the halo), and computationally because the size of the each chunk that is sent to processing is approximately the same, which therefore optimizes the execution of the code. In our particular case, we are using HiPSCat pixels that are a few hundred megabytes in size.

\subsection{Pipeline Performance}

We have performed several benchmarks to estimate the performance of the PhotoD code. In our initial tests based on the SDSS Stripe 82 data we found that the distances for a billion stars can be estimated in about 100 days on a Mac M1 Pro laptop using one core (10 milliseconds per star).

In further testing, we used outputs from the TRILEGAL simulation and found that the parallel processing performed with LSDB scaled well with respect to the single core values. On a local cluster node with dual AMD EPYC 9474F 48-core processors we were able to achieve per-star figures similar to single-core processing, therefore reducing the required processing time to about a week. Our ongoing tests are showing that when all 10 nodes (400 cores) of the Ruđer Bošković Institute's Narval cluster are used, we may be able to estimate distances for about 10 billion stars in about two days.

The output catalog containing LSST photometry (with errors) and minimalistic auxiliary metadata for 10 billion stars are would amount to about 1 TB of data. The code outputs, including a covariance matrix of the three fitted parameters, would have approximately the same volume.
